Argennon: A Scalable Cloud Based Smart Contract Platform Using Argument of Knowledge Systems

### 2023-01-11 tex/sections/avm.tex
Improve readability of Chapter 2
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

The Argennon Smart Contract Execution Environment (AscEE) is an abstract high level execution environment for executing
Argennon smarts contracts (a.k.a\ Argennon applications) in an efficient and isolated environment. An Argennon
application essentially is an HTTP server whose state is
kept in the Argennon blockchain and its logic is described using an Argennon Standard Application Representation (ASAR).

@@

The state of an Argennon application is stored in byte addressable finite arrays of memory called
\emph{heap chunks}. An application may have several heap chunks with different sizes, and can remove or
resize its heap chunks or allocate new chunks. Every chunk belongs to exactly one application and can only be modified
by its owner. In addition to heap chunks, every application has a limited amount of non-persistent local memory for
storing temporary data.

The AscEE executes the requests contained in each block of the Argennon
blockchain in a three-step procedure. The first step is the \emph{preprocessing step}. In this step, the required
data for executing requests are retrieved and verified and the
helper data structures for next steps are constructed. This step is
designed in a way that can be done fully in parallel for each request without any risk of data races. The second
step is the \emph{Data Dependency Analysis (DDA) step}.
In this step by analyzing data dependency between requests, the AscEE determines requests that can be run in parallel
and requests that need to be run sequentially. This information is represented using an \emph{execution DAG} data
structure and in the final step, the requests are executed using this data structure.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing Argennon applications to
fulfill \emph{external} HTTP requests\footnote{External requests are requests that are not made by other Argennon
applications.}, produce their HTTP responses and update related heap chunks. The execution of requests can be
considered sequential\footnote{Actually requests may be executed in parallel but by performing data dependency analysis
the result is guaranteed to be identical with the sequential execution of requests.} and each request has a separate
\emph{execution session}. In other words, an execution session is a separate session of executing application's code in
order to fulfill an external HTTP request. We call this external request the \emph{initiator} of the execution session.

@@

Every Argennon application has two types of memory: local memory and heap. Local memory is not persistent and is
destroyed when the application call ends. Local memory is used for storing local variables and is not directly
addressable. Heap, on the other hand, is persistent and can be used for persisting data between application calls.
Heap is addressable and provides low level direct access. Both local memory and heap are limited, but the
AscEE does not specify any particular limits for them. When an application tries to use too much memory, that may
cause the execution session to end abruptly. In that case, the execution session will not have any effects on the
state of heap.

@@

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself, provided the new id has a correct format. This
is an important feature of the AscEE heap which allows applications to use the AscEE heap as a map
data structure\footnote{also called a dictionary.}.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily find unique identifiers for its chunks.

During an execution session every heap chunk has an access-type which may disallow certain operations. The
access-type of a chunk is declared by the initiator request of the execution session:

\begin{itemize}
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{additive}: only allows additive operations. By additive we mean an addition-like operator without
    overflow checking which is associative and commutative. Note that the content of these chunks cannot be read.
\end{itemize}

@@

At the start of executing requests of a block, a validator can consider two values for every
heap chunk, the size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at the start of
every execution session, and it may be updated during the session by the owner application. On the other hand,
the value of \texttt{sizeUpperBound} is proposed by the block proposer for each block and is calculated based on values
declared by external requests (i.e.\ transactions) that want to perform chunk resizing. This value needs to be
an upper bound of all the declared resizing values and indicates the upper bound of \texttt{chunkSize} during the
execution of a block.

The address space of a chunk starts from zero and only offsets lower than \texttt{sizeUpperBound} are valid. Trying to
access any offset higher than \texttt{sizeUpperBound} will always cause the execution session to end abruptly. There
is no way for an application to query \texttt{sizeUpperBound}. As a result, in the view of an application,
accessing offsets higher than \texttt{chunkSize} will result in undefined behaviour, while the behaviour is
well-defined for validators.
This enables validators to determine the validity of an offset at the start of the block validation in a parallelized
preprocessing phase without actually executing requests.

The value of \texttt{chunkSize}, can be modified during an execution session. However, the new values of size can
only be increasing or decreasing. More precisely, if a request declares that it wants to expand (shrink) a chunk, it
can only increase (decrease) the value of \texttt{chunkSize} and any specified value during the execution
session, needs to be greater (smaller) than the previous value of the \texttt{chunkSize}. Any request that wants to
expand (shrink) a chunk needs to specify a max size (min size). The value of \texttt{chunkSize} can not be set higher
(lower) than this value.

@@

An application can load any chunk with a valid prefix identifier even if that chunk does not exist. For a non-existent
chunk the value of \texttt{chunkSize} is always zero.

@@

\begin{itemize}
    \item \texttt{appID} is a prefix code built by \emph{applications} prefix tree. An \texttt{appID} cannot
    be \texttt{0x0}.

    \item \texttt{accountID} is a prefix code built by \emph{accounts} prefix tree. An \texttt{accountID} cannot
    be \texttt{0x0} or \texttt{0x1}.

    \item \texttt{chunkID} is a composite prefix code built by concatenating an \texttt{applicationID} to
    an \texttt{accountID} to a prefix code made by \emph{local} prefix tree:
    \subitem \texttt{chunkID = (applicationID|accountID|<local-prefix-code>)} .
\end{itemize}

@@

When we have a prefixed identifier, and we want to know if a sequence of digits is marked by that identifier,
we use Algorithm~\ref{alg:match_id} to match the prefixed identifier with the start of the sequence. The matching
can be done with only three comparisons, and an invalid prefixed identifier can be detected and will not match
any sequence.

@@

The attachment of a request is a list of request identifiers of the current block that are \emph{attached} to the
request. That means, for validating the request a validator first needs to \emph{inject} the digest of attachments
into the current HTTP request text. By doing so, the called application will have access to the digest of
attachments in a secure and authenticated way.

The main usage of this feature is for fee payment. A request that wants to pay the fees for a number of requests,
declares those requests as its attachments. For paying fees the payer signs the digest of requests for which he
wants to pay fees. After injecting the digest of those request by validators, that signature can be validated
correctly and securely by the application that handles fee payment.

>>>>

The Argennon Smart Contract Execution Environment (AscEE) is an abstract high level execution environment for executing
Argennon smarts contracts a.k.a\ Argennon applications in an efficient and isolated environment. An Argennon
application essentially is an HTTP server whose state is
kept in the Argennon blockchain and its logic is described using an Argennon Standard Application Representation (ASAR).

@@

The state of an Argennon application is stored in byte addressable finite arrays of memory called
\emph{heap chunks}. An application may have several heap chunks with different sizes, and can remove or
resize its heap chunks or allocate new chunks. Every chunk belongs to exactly one application and can only be modified
by its owner. Besides heap chunks, every application has a limited amount of non-persistent local memory for
storing temporary data.

The AscEE executes the requests contained in each block of the Argennon
blockchain in a three-step procedure. The first step is the \emph{preprocessing step}. In this step, the required
data for executing requests are retrieved and verified and the
helper data structures for next steps are constructed. This step is
designed in a way that can be done fully in parallel for each request without any risk of data races. The second
step is the \emph{Data Dependency Analysis (DDA) step}.
In this step by analyzing data dependency between requests, the AscEE determines requests that can be run in parallel
and requests that need to be run sequentially. This information is represented using an \emph{execution DAG} and in
the final step, the requests are executed using this data structure.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing Argennon applications to
fulfill \emph{external} HTTP requests\footnote{External requests are requests that are not made by other Argennon
applications.}, produce their HTTP responses and update related heap chunks. The execution of requests can be
considered sequential\footnote{Actually requests may be executed in parallel but by performing data dependency analysis
the result is guaranteed to be identical with the sequential execution of requests.} and each request has a separate
\emph{execution session}. An execution session is a separate session of executing application's code in
order to fulfill an external HTTP request. This external request is the \emph{initiator} of the execution session.

@@

Every Argennon application has two types of memory: local memory and heap. Local memory is not persistent and is
destroyed when the application call ends. Heap, on the other hand, is persistent and can be used for persisting data
between application calls. Local memory is used for storing local variables and is not directly
addressable. Heap is addressable and provides low level direct access. Both local memory and heap are limited, but
the limit is not specified by the AscEE. If an application tries to use too much
memory, that may cause the execution session to end abruptly. In that case, the execution session will not have any
effects on the state of heap.

@@

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself, provided the new identifier has a correct format.
This is an important feature of the AscEE heap which enables applications to use the AscEE heap as a map
data structure\footnote{also called a dictionary.}.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily find unique identifiers for its chunks. (See Section~\ref{sec:identifiers})

During an execution session every heap chunk has an access-type which may disallow certain operations during that
session. This access-type is declared by the initiator request of the execution session for every chunk:

\begin{itemize}
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{additive}: only allows additive operations. By additive, we mean an addition-like operator without
    overflow checking which is associative and commutative. Note that the content of the chunk cannot be read.
\end{itemize}

@@

At the start of executing requests of a block, a validator can consider two values for every
heap chunk, the size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at the start of
every execution session, and it may be updated during the session by the owner application. On the other hand,
the value of \texttt{sizeUpperBound} can be determined uniquely at the start of block validation and is proposed by the
block proposer for each block. This value is calculated based on resizing values
declared by external requests (i.e.\ transactions) that want to perform chunk resizing and needs to be
an upper bound of all the declared resizing values, indicating an upper bound of \texttt{chunkSize} during the
execution of the requests of a block.

The address space of a chunk starts from zero and only offsets lower than \texttt{sizeUpperBound} are valid. Trying to
access any offset higher than \texttt{sizeUpperBound} will always cause the execution session to end abruptly. The
value of \texttt{sizeUpperBound} is always greater than \texttt{chunkSize} and there is no way for an application to
query \texttt{sizeUpperBound}. As a result, in the view of an application,
accessing offsets higher than \texttt{chunkSize} results in undefined behaviour, while the behaviour is
well-defined in the view of a validator.
This enables validators to determine the validity of an offset at the start of the block validation in a
preprocessing phase without actually executing requests, while \texttt{sizeUpperBound} can be determined in a simple
parallelized algorithm.

The value of \texttt{chunkSize}, can be modified during an execution session. However, the new values of size can
only be increasing or decreasing. More precisely, if a request declares that it wants to expand (shrink) a chunk, it
can only increase (decrease) the value of \texttt{chunkSize} and any specified value during the execution
session, needs to be greater (smaller) than the previous value of \texttt{chunkSize}. Any request that wants to
expand (shrink) a chunk needs to specify a max size (min size). The value of \texttt{chunkSize} can not be set higher
(lower) than this value.

@@

An application may load any chunk with a valid prefix identifier even if that chunk does not exist. For a non-existent
chunk the value of \texttt{chunkSize} is always zero.

@@

\begin{itemize}
    \item \texttt{appID} is a prefix code built by the \emph{applications} prefix tree. An \texttt{appID} cannot
    be \texttt{0x0}.

    \item \texttt{accountID} is a prefix code built by the \emph{accounts} prefix tree. An \texttt{accountID} cannot
    be \texttt{0x0} or \texttt{0x1}.

    \item \texttt{chunkID} is a composite prefix code built by concatenating an \texttt{applicationID} to
    an \texttt{accountID} to a prefix code made by the \emph{local} prefix tree:
    \subitem \texttt{chunkID = (applicationID|accountID|<local-prefix-code>)} .
\end{itemize}

@@

When we have a prefixed identifier, and we want to know if a sequence of digits is marked by that identifier,
we use Algorithm~\ref{alg:match_id} to match the prefixed identifier with the start of the sequence. The matching
can be done with only three comparisons, while invalid identifiers can be detected and will not match
any sequence.

@@

The attachment of a request is a list of request identifiers of the current block that are \emph{attached} to the
request. That means, for validating this request a validator first needs to \emph{inject} the digest of attachments
into its HTTP request text. By doing so, the called application will have access to the digest of
attachments in a secure way, while it is ensured that the attached requests are included in the
current block.

The main usage of this feature is for fee payment. A request that wants to pay the fees for a number of requests,
declares those requests as its attachments. For paying fees the payer signs the digest of requests for which he
wants to pay fees. After injecting the digest of those request by validators, that signature can be validated by the
application that handles fee payment, and it is guaranteed that the attached requests are actually included in the
current block.
